A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-01-08 analysis/paper/paper.Rmd
Fixed unpacking of results

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @berg-hansen2009; @bjerck2005; @sognnes2003; @tallavaara2020].

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from a resource perspective and from one of movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes. This general picture is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, where the introduction of the Neolithic proper inolved a. associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@solheim2018; @prescott2012]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods.

@@

One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether they are entirely erroneous results of contamination can be difficult to evaluate. Here this will be entirely dependent on, and follow the discretion of the original excavation reports.

>>>>

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @berg-hansen2009; @bjerck2005; @sognnes2003; @solheim2018; @tallavaara2020].

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper. associated involving a change agro-pastoralism, the farm,  means that the site locations are now more withdrawn from the shoreline [@solheim2018; @prescott2012]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods. Earlier, at the transition to the Early Neolithic (c. 4000 BCE) pottery is introduced to the sites, and there are some possible indications of an initial, small-scale uptake of agriculture at some sites in the Oslo fjord region. However, this initial uptake appears to be relatively small-scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, followed by a de-Neolithisation in the Middle Neolithic. [@nielsen2021] has recently argued that while there, the initial uptake of in the Early Neolithic is combined with a more complex settlement pattern, and argues that the simple forager/farming dichotomy has underplayed the variation present in the Early and Middle Neolithic data. The empirical expectation for this development would thus be a more varied relation between sites and the shore-line.

@@

One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues (@breivik2018; @solheim2020), who compared 102 ^14^C-dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate. Here this will be entirely dependent on, and follow the discretion of the original excavation reports.
